A Computational Account of Default Mode Function\\ by Reinforcement Learning Theory

### 2017-03-26 paper/paper.tex
integrated Elvis' comments

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
In the brain, such an experience replay mechanism is probably implemented in the hyppocampus (HC). Indeed,
Hippocampal replay is a phenomenon observed in rats, mice, cats, rabbits, songbirds, and monkeys. Indeed, during sleep or awake rest, replay refers to the re-occurrence of a sequence of cell activations that also occurred during activity, but the replay has a much faster time scale.
% It may be in the same order, or in reverse.
Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For example, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action $(s,a)$ into an appropriate
feature space. Such are the techniques that have been proposed by \cite{songNIPS2016}, for example.
%  and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind \cite{mnih2015,silver2016mastering}, is to learn this
representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which
are by far the current state-of-the-art in RL. In such a case, $\theta^Q$
would correspond to the strengths of low-level synaptic or high-level functional connections between different
neurons / nodes / regions in the network.

@@

\section{Relation to other statistical models (Elvis)}
\subsection{Friston's Free-energy principle (refer to \ref{label:friston})}
\textbf{XXX: add tons of more examples and connnections}
In summary:
\begin{itemize}
  \item Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute with the free energy principle
  in practice, which makes it physiologically less plausible. Auto-encoder theory presented
  \ref{sec:vae} can help to remove these barriers, by making the whole scheme trainable via
  backprop!
  \end{itemize}

@@

\begin{itemize}
  \item Both the predictive coding account and MDPs
  are process theories backed up by plausible
  neurobiological evidence that
  view the brain as a ``statistical organ''
  generalizing from the past to new experiences.
  \item Both provide a parsimonious explanation why the
  human brain decreases processing load devoted to incoming information
  when the environment is predictable and allocate increased
  attentional resources when novel stimuli are encountered.
  Similarly, both propose explanations why
  predicting the future is inherently
  linked to information from the past.
  \item Sensory experience is a generative process from both views.
  In predictive coding, sensory perception of the external world
  is a generative experience due to modulatory top down experience at
  various hierarchical levels of sensory input processing.
  In our RL view, the (partially observed)
  environmental model is incorporated in the MDP,
  \textit{which can be fully recovered based on the last
  state alone}.
  \item Both naturally expose a mechanisms of brain plasticity in that
  neuronal wiring gets increasingly adapted
  when faced by unprecedented environmental
  challenges.
  \item The hierarchical aspects from predictive coding
  is re-expressed in MDPs in form of
  nested prediction of probable upcoming actions and rewards.
  \item Both model the consequences of action. In RL, the horizon of that
  look into the future is explicitly manifested in the $\gamma$ parameter
  in the Bellman equation, but not
  explicitly modeled in the PC account that
  tends to emphasize prediction about the
  immediate future.
  \item Mismatch negativity of PC is immanent in reward prediction error
  by the difference between actually predicted reward of an action given
  a state and the reward predicted by non-linear regression with graded
  discounting of future rewards.
  \item Adapting the neuronal connections for improved top-down modulation
  takes the concrete form of gradient computation
  \textit{and back-propagation} in MDPs and RL,
  although this may not be neurobiologically plausible.
  \item In sum,
  the MDP account may serve as a concrete conceptualization of the PC account
  in cognitive neuroscience. MPDs have the advantage of exposing an explicit
  mechanism of the horizon of future considerations or
  how the internal model of the world gets updated,
  and collapses sensory input processing and action output preparation.
  %
  As such, our account can be viewed as a RL implemnetion of the
  prediction coding hypothesis.
\end{itemize}

\subsection{Semantic Hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extractable from the environment
\citep{bar2009proactive}.
%
Going much beyond human language ifself,
semantic building blocks provide the basis to
mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
a constructive account explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
Cognitive aspects of egocentric-allocentric switching
are also closely related to episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\citep{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \citep{tolman1948cognitive}.

>>>>

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
In the brain, such an experience replay mechanism is probably implemented in the hyppocampus (HC). Indeed,
Hippocampal replay is a phenomenon observed in rats, mice, cats, rabbits, songbirds, and monkeys. Indeed, during sleep or awake rest, replay refers to the re-occurrence of a sequence of cell activations that also occurred during activity, but the replay has a much faster time scale.
% It may be in the same order, or in reverse.
Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For example, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action $(s,a)$ into an appropriate
feature space. Such are the techniques that have been proposed by [sutton et al. xyzw, Song et al. 2016 (NIPS),
  and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind [Minh et al. 2014, Silver et al 2015] is to learn this
representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which
are by far the current state-of-the-art in RL. In such a case, $\theta^Q$
would correspond to the strengths of low-level synaptic or high-level functional connections between different
neurons / nodes / regions in the network.

@@

\section{Relation to other statistical models (Elvis)}
\subsection{Friston's Free-energy principle (refer to \ref{label:friston})}
In summary:
\begin{itemize}
  \item Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute with the free energy principle
  in practice, which makes it physiologically less plausible. Auto-encoder theory presented
  \ref{sec:vae} can help to remove these barriers, by making the whole scheme trainable via
  backprop!
  \end{itemize}

@@

\begin{itemize}
  \item Both the predictive coding account and MDPs
  are process theories backed up by plausible
  neurobiological evidence that
  view the brain as a ``statistical organ''
  generalizing from the past to new experiences.
  \item Both provide a parsimonious explanation why the
  human brain decreases processing load devoted to incoming information
  when the environment is predictable and allocate increased
  attentional resources when novel stimuli are encountered.
  Similarly, both propose explanations why
  predicting the future is inherently
  linked to information from the past.
  \item Sensory experience is a generative process from both views.
  In predictive coding, sensory perception of the external world
  is a generative experience due to modulatory top down experience at
  various hierarchical levels of sensory input processing.
  In our RL view, the (partially observed)
  environmental model is incorporated in the MDP,
  \textit{which can be fully recovered based on the last
  state alone}.
  \item Both naturally expose a mechanisms of brain plasticity in that
  neuronal wiring gets increasingly adapted
  when faced by unprecedented environmental
  challenges.
  \item The hierarchical aspects from predictive coding
  is re-expressed in MDPs in form of
  nested prediction of probable upcoming actions and rewards.
  \item Both model the consequences of action. In RL, the horizon of that
  look into the future is explicitly manifested in the $\gamma$ parameter
  in the Bellman equation, but not
  explicitly modeled in the PC account that
  tends to emphasize prediction about the
  immediate future.
  \item Mismatch negativity of PC is immanent in reward prediction error
  by the difference between actually predicted reward of an action given
  a state and the reward predicted by non-linear regression with graded
  discounting of future rewards.
  \item Adapting the neuronal connections for improved top-down modulation
  takes the concrete form of gradient computation
  \textit{and back-propagation} in MDPs and RL,
  although this may not be neurobiologically plausible.
  \item In sum,
  the MDP account may serve as a concrete conceptualization of the PC account in cognitive neuroscience. MPDs have the advantage of exposing an explicit
  mechanism of the horizon of future considerations or
  how the internal model of the world gets updated,
  and collapses sensory input processing and action output preparation.
\end{itemize}

\begin{mdframed}
  \textbf{XXX:} Rather than a heads-on competitor, our proposed formalism should be seen
  and sold as a solid implementation of the predictive coding hypothesis.
\end{mdframed}
\subsection{Semantic Hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extractable from the environment
\citep{bar2009proactive}.
%
Going much beyond human language ifself,
semantic building blocks provide the basis to mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
a constructive account explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
Cognitive aspects of egocentric-allocentric switching
are also closely related to episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\citep{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \citep{tolman1948cognitive}.
